Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-09-18 Manuscript/manuscriptPS.tex
Update text

\emph{NMR numerical simulations}
The numerical simulations of S-DROSS curves were performed with the SIMPSON simulation package using the $^{13}$C--$^1$H
dipolar couplings determined by the R-PDLF experiments or calculated from known $^2$H quadrupolar couplings as input.
Chemical shift anisotropy and homonuclear couplings were neglected, and the input file {\it{rep2000}} was used to simulate the random
distribution of bilayer orientations in the samples studied.

@@

The C-H bond order parameters were calculated directly
from the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\langle 3\cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C-H bond and the membrane normal.
Angular brackets point to the average over all sampled configurations.
The order parameters were first calculated separately for each lipid molecule
in the system averaging over time. The average and
the standard error of the mean were then calculated over different lipids.
The number density profiles were calculated using {\it gmx density} tool
from Gromacs sofware package \cite{gromacsMANUAL}.

@@

In the experiment from the literature used in this work~\cite{browning80,roux90},
the lipids were first soluted to the buffer and then centrifuged to a pellet which
was used in the measurements. In such experiments the samples contain approximately
10~wt~\% of lipids~\cite{browning80,roux88,roux90}, which is less than in
gravimetric experiments (60~wt~\%) and simulations (approx. 50-60~wt~\%) in this work.
Since multilamellar structures of pure PS lipids swell due to the electrostatic repulsion,
the larger repeat distances are expected in samples with lower lipid concentrations~\cite{millman82}.
However, the PS headgroup order parameters from gravimetric sample are in good agreement
with centrifuged sample (Fig. \ref{HGorderParameters}) and the equlibrium repeat distance decreases
already above 500~mM concentration of monovalent ions close to the simulation box sizes \cite{millman82,rand89}.
Therefore, the hydration levels of multilamellae are expected to be sufficiently similar
in simulations and the reference experiments.

Two different definitions for the salt concentrations have been used when
electrometer concept is applied to study ion binding affinity.
The concentrations are reported either in water before solvating the lipids \cite{akutsu81,roux90,catte16}
or in bulk water after solvating the lipids \cite{altenbach84,melcr18}.
In this work, we use the former definition to be consistent with the reference
experimental data \cite{roux90}. The used definition has only a marginal effect
to the results is simulations with realistic ion binding affinity.

@@

\todo{How is the peak assignment done?}

The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectras for all the
carbons in headgroup and glycerol backbone region, except for g$_3$ for which the resolution
was not sufficient to determine the numerical value of the order paramater (Fig. \ref{PShgSIGNSsimpson}).
Slices of the R-PDFL spectra (Fig. \ref{PShgSIGNSsimpson} C)) 
show a single splitting for the $\beta$-carbon with the order parameter value of 0.12,
and a superposition of a large and a very small splitting for the $\alpha$-carbon.
The larger splitting gives a order parameter value of 0.09, while the numerical value
from the small splitting cannot resolved with the available resolution.
Since the R-PDFL and previous $^2$H NMR experiments \cite{browning80,roux91} give 
only the absolute values of order parameters, we determined the signs of PS headgroup
and glycerol backbone order parameters using the S-DROSS experiment \cite{gross97}.
The S-DROSS slice for the $\beta$-carbon (Fig. \ref{PShgSIGNSsimpson} D)) clearly shows that the
order parameter is negative, which is confirmed by SIMPSON simulations.
The beginning of the S-DROSS slice suggests that the higher order parameter of
the $\alpha$-carbon is positive and the deviation towards negative values with the longer T$_1$ times suggests
that the smaller order parameter is negative. This is confirmed by a SIMPSON simulation
where the value of -0.02 was taken from $^2$H NMR experiment \cite{roux91} for the smaller order parameter.
The literature value was used because the
resolution of our experiment was not sufficient to determine the
small value of the order parameter. The S-DROSS curve from
SIMPSON simulation with a positive value for the smaller order parameter
(dashed grey in Fig. \ref{PShgSIGNSsimpson} D)) did not agree with the experiment, 
confirming the interpretation that the smaller order parameter is negative.
\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
  \caption{\label{PShgSIGNSsimpson}
    The headgroup and glycerol backbone region of the (A) INEPT spectrum and
    (B) 2D R-PDLF spectra.
    (C) Experimental SDROSS data (points) and SIMPSON simulations (blue lines) with
    the order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
    for the $\alpha$-carbon slittings. The S-DROSS curve from SIMPSON simulation with positive value
    for the smaller $\alpha$-carbon order parameter (dashed grey).
  }
\end{figure}

%\begin{figure}[!htb]
%  \centering
%  \includegraphics[width=5.5cm]{../Figs/g1_slice_large.eps}
%  \includegraphics[width=5.5cm]{../Figs/g1_slice_small.eps}
%  \includegraphics[width=5.5cm]{../Figs/g2_slice.eps}
%  \caption{\label{glycerolNMR}
%    R-PDFL slices for gycerol backbone carbons.
%  }
%  \todo{We need nicer figure for this data. Maybe combine with \ref{PShgSIGNSsimpson}} \\
%  \todo{What are the top figures actually?}
%\end{figure}

@@

To evaluate counterion binding in different simulation models against experimental data~\cite{roux90},
we plot the headgroup order parameters measured from POPC:POPS 5:1 mixture
as a function of different monovalent ions added to the buffer (Fig. \ref{PSresponseTONaCl}). 
Experimental order parameter data for POPC headgroup in the mixture is available as a function
of LiCl and KCl concentrations, while POPS headgroup order parameters are measured also
as a function of NaCl. Lithium interacts more strongly with PS headgroups than other monovalent 
ions~\cite{hauser83,hauser85,roux86,mattai89,roux90}, as also observed for PC headgroups~\cite{cevc90}. 
This is evident also in the changes of PS headgroup order parameters, which decrease with the addition of lithium 
but increase with the addition of sodium or potassium (Fig. \ref{PSresponseTONaCl}). 
POPC headgroup order parameters exhibit a clear decrease as a function of LiCl concentration
but only modest changes as a function of KCl concentration, indicating singificant 
Li$^+$ binding but only weak Na$^+$ binding to the mixture when interpreted using the
electrometer concept~\cite{akutsu81,altenbach84,seelig87}. In simulations with the
Berger model, the headgroup order parameter response of POPC
to the added NaCl is similar to the experiments of LiCl,
indicating overestimated binding affinity of sodium, in line with the results for PC bilayers \cite{catte16}.
Indeed, the sodium density profile shows a significant binding peak in the 
Berger model (Fig. \ref{CIdensPSOCmixt}). Potassium binding in the MacRog simulation
is significantly weaker  (Fig. \ref{CIdensPSOCmixt}) and the headgroup order parameter 
changes are also in better agreement with simulations (Fig. \ref{PSresponseTONaCl}).
\todo{Discussion about Lipid17 to be written when we have the density profiles.}
All the tested models overestimate the changes of POPS headgroup order parameters as
a function of monovalent ions (Fig. \ref{PSresponseTONaCl}), suggesting that
model development is necessary to interpret the PS headgroup-ion interactions 
from MD simulations.
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
  \caption{\label{PSresponseTONaCl}
    Changes of the PC (left) and PS (right) headgroup order parameters as a function of
    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture at 298~K
    (except Berger simulations are (4:1) mixture at 310~K).
    The experimental data is from Ref. \citenum{roux90}.
    The values from counterion-only systems are set as a zero point of y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
    parameter value is at zero.
  }
  \todo{CHARMM36 results for this plot would be highly useful.}
\end{figure*}

>>>>

\emph{NMR numerical simulations}
The numerical simulations of S-DROSS curves were performed with the SIMPSON simulation package \cite{bak00} using the $^{13}$C--$^1$H
dipolar couplings determined by the R-PDLF experiments or calculated from the known $^2$H quadrupolar couplings \cite{browning80} as input.
Chemical shift anisotropy and homonuclear couplings were neglected, and the input file {\it{rep2000}} was used to simulate the random
distribution of bilayer orientations in the samples studied.

@@

The C-H bond order parameters were calculated directly
from the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\langle 3\cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C-H bond and the membrane normal.
Angular brackets point to the average over all sampled configurations.
The order parameters were first calculated averaging over time separately
for each lipid molecule in the system. The average and
the standard error of the mean were then calculated over different lipids.
The number density profiles were calculated using {\it gmx density} tool
from Gromacs sofware package \cite{gromacsMANUAL}.

@@

In the experimental $^2$H NMR literature data used in this work~\cite{browning80,roux90},
the lipids were first soluted to the buffer and then centrifuged to a pellet which
was used in the measurements. Such samples have lower lipid concentration
(approximately 10~wt~\% of lipids~\cite{browning80,roux88,roux90}) than 
gravimetric samples (60~wt~\%) and simulations (approximately 50-60~wt~\%) in this work.
Larger multilamellar repeat distances are expected in the samples with lower lipid
concentrations due to the swelling caused by electrostatic repulsion in pure PS lipid systems~\cite{millman82}.
However, the PS headgroup order parameters measured from gravimetric sample in this work
are in good agreement with the results from centrifuged sample in the literature \cite{browning80} (Fig. \ref{HGorderParameters}).
Furthermore, the equlibrium repeat distance rapidly decreases with the addtion of monovalent
salts and is close to the simulation box sizes already above 500~mM concentrations~\cite{millman82,rand89}.
Therefore, the hydration levels of multilamellae are expected to be sufficiently similar
in the used simulations and reference experiments.

Two different definitions for the salt concentrations have been used when
electrometer concept is applied to study ion binding affinity.
The concentrations are reported either in water before solvating the lipids \cite{akutsu81,roux90,catte16}
or in bulk water after solvating the lipids \cite{altenbach84,melcr18}.
In this work, we use the former definition to be consistent with the reference
experimental data \cite{roux90}. The used definition has only a marginal effect
to the results is simulations with realistic ion binding affinity.
\todo{Maybe we should add a figure in the SI.}

@@

The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectras for all the
carbons in headgroup and glycerol backbone region (Fig. \ref{PShgSIGNSsimpson}).
The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13}.
The peaks for $\beta$ and $\alpha$ carbons were assigned according to the
known order parameters from the $^2$H NMR experiments~\cite{browning80}.
Slices of the R-PDFL spectra and the resulting order parameters values
are shown in the supplementary information (Fig. \ref{DPslices}). 
Since the R-PDFL and previous $^2$H NMR experiments \cite{browning80,roux91} give 
only the absolute values of order parameters, we determined the signs of PS headgroup
order parameters using the S-DROSS experiment \cite{gross97}.
The S-DROSS slice clearly shows that the order parameter of
the $\beta$-carbon is negative (Fig. \ref{PShgSIGNSsimpson} C)),
which is confirmed by SIMPSON simulations. The beginning of the S-DROSS slice
suggests that the larger order parameter of the $\alpha$-carbon 
is positive and the deviation towards negative values with longer T$_1$ times suggests
that the smaller order parameter is negative. This is confirmed by a SIMPSON simulation
using the value of -0.02 from $^2$H NMR experiment \cite{roux91} for the smaller order parameter.
The literature value was used because the
resolution of our experiment was not sufficient to determine the
small value of the order parameter.
The S-DROSS curve from SIMPSON simulation with a positive value for the smaller order parameter
(dashed grey in Fig. \ref{PShgSIGNSsimpson} C)) did not agree with the experiment, 
confirming the interpretation that the smaller order parameter is negative.
\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
  \caption{\label{PShgSIGNSsimpson}
    The headgroup and glycerol backbone region of the (A) INEPT spectrum and
    (B) 2D R-PDLF spectra.
    (C) Experimental SDROSS data (points) and SIMPSON simulations (blue lines) with
    the order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
    for the $\alpha$-carbon slittings. The S-DROSS curve from SIMPSON simulation with positive value
    for the smaller $\alpha$-carbon order parameter (dashed grey).
  }
\end{figure}

@@

To evaluate counterion binding in different simulation models against experimental data~\cite{roux90},
we plot the headgroup order parameters measured from POPC:POPS 5:1 mixture
as a function of different monovalent ions added to the buffer (Fig. \ref{PSresponseTONaCl}). 
Experimental order parameter data for POPC headgroup in the mixture is available as a function
of LiCl and KCl concentrations, while POPS headgroup order parameters are measured also
as a function of NaCl. Lithium interacts more strongly with PS headgroups than other monovalent 
ions~\cite{hauser83,hauser85,roux86,mattai89,roux90}, as also observed for PC headgroups~\cite{cevc90}. 
This is evident also in the changes of PS headgroup order parameters, which decrease with the addition of lithium 
but increase with the addition of sodium or potassium (Fig. \ref{PSresponseTONaCl}). 
POPC headgroup order parameters exhibit a clear decrease as a function of LiCl concentration
but only modest changes as a function of KCl concentration, indicating singificant 
Li$^+$ binding but only weak Na$^+$ binding to the mixture when interpreted using the
electrometer concept~\cite{akutsu81,altenbach84,seelig87}. In simulations with the
Berger model, the headgroup order parameter response of POPC
to the added NaCl is similar to the experiments of LiCl,
indicating overestimated binding affinity of sodium, in line with the results for PC bilayers \cite{catte16}.
Indeed, the sodium density profile shows a significant binding peak in the 
Berger model (Fig. \ref{CIdensPSOCmixt}). Potassium binding in the MacRog simulation
is significantly weaker  (Fig. \ref{CIdensPSOCmixt}) and the headgroup order parameter 
changes are also in better agreement with simulations (Fig. \ref{PSresponseTONaCl}).
\todo{Discussion about Lipid17 to be written when we have the density profiles.}
All the tested models overestimate the changes of POPS headgroup order parameters as
a function of monovalent ions (Fig. \ref{PSresponseTONaCl}), suggesting that
model development is necessary to interpret the PS headgroup-ion interactions 
from MD simulations.
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
  \caption{\label{PSresponseTONaCl}
    Changes of the PC (left) and PS (right) headgroup order parameters as a function of
    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture at 298~K
    (except Berger simulations are (4:1) mixture at 310~K).
    The experimental data is from Ref. \citenum{roux90}.
    The values from counterion-only systems are set as a zero point of y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
    parameter value is at zero.
  }
\end{figure*}

### 2018-09-18 Manuscript/manuscriptPSsuppl.tex
Remove POPS vs OPPS discussion from the SI
Update text

\pagebreak
\section{Dipolar slices of the R-PDFL experiment}
\begin{figure}[]
%  \centering
  \includegraphics[width=20.0cm]{../Figs/SI_slices.pdf}
  \caption{\label{DPslices}
    Dipolar coupling slices from the R-PDFL experiment.
  }
  \todo{Labeling of carbons and explanations of numbers (chemical shift and order parameter) would be good.
  Also, the size of the figure in the file may not be optimal.}
\end{figure}

@@

\pagebreak
\section{Difference between POPC and OPPS in MacRog model}

\todo{This is to be removed once the new data is available: \\
https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/16}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPOPSvsOPPS.eps}
  \caption{\label{CHANGESwithCaClPGPS}
    Headgroup order parameters from POPS and OPPS simulations with MacRog model.}
\end{figure}

>>>>

\pagebreak
\section{Dipolar slices of the R-PDFL experiment}

Slices of the R-PDFL spectra (Fig. \ref{DPslices}) 
show a single splitting for the $\beta$-carbon with the order parameter value of 0.12,
and a superposition of a large and a very small splitting for the $\alpha$-carbon.
The larger splitting gives a order parameter value of 0.09, while the numerical value
from the small splitting cannot resolved with the available resolution.
\begin{figure}[]
%  \centering
  \includegraphics[width=20.0cm]{../Figs/SI_slices.pdf}
  \caption{\label{DPslices}
    Dipolar coupling slices from the R-PDFL experiment.
  }
  \todo{Labeling of carbons and explanations of numbers (chemical shift and order parameter) would be good.
  Also, the size of the figure in the file may not be optimal.}
\end{figure}

@@
